Ambient: overview of the workflow and design basics

### 2011-09-19 doc/paper_ambient/paper_ambient.tex
paper

In the last decade, the distributed memory were confirmed like the technology in HPC architecture machine, 100 \% of the top 500 are now based on this model, as the large scalability applications
used  a main MPI layer, we can cite as popular   in a few scientist field :  [GIVE ME A MPI CODE INTO ALPS !!!!]   {\sc Quantum ESPRESSO} \cite{QE-2009} in quantum chemistry, Abinit \cite{ABINIT-2005, ABINIT-2009} in 
quantum mechanic,   ScaLAPACK \cite{SCALAPACK-1997} in linear  algebra or SMILE in hypersonic rarefied flow \cite{SMILE-2005}. Of course, this short list is not exhaustive, and it can be fill up easily in every fields of 
science and engineering. The limitation of this programming model (the scalability) will be the communication overhead and the workload imbalance among process.

Secondly, the multicore technology  introduced by the ship makers the last decade.  Indeed, the progress of the industrial process allowed the integration of several cores into the same die. 
This technology is now available on all  processor architecture (Power, X86, ARM, \etc). One more time, the programming 
model was modified, and it re-actualized the shared memory model.  To program shared memory machine, a proved technology exists, it is the POSIX threads library. 
The programmers have two possibilities. They utilize the library directly (pThreads) or he lets the compiler generates a multi-thread executables.  
Generally, the former option places the burden on the programmer, but when well done, it provides good performance.
On the other hand,  the second option more knew under the  name OpenMP  \cite{openmp:2002}, easier,  is based on  preprocessor directive. This second option
is clearly the most popular nowadays.

As MPI, OpenMP is developed into a consortium, the version 1.0 was released in 1997, the 2.0 in 2005 and new features were integrated in 2008.
This programming model becomes popular, it is now integrated natively in all compilers and it allows hybrid programs. Hybrid mode MPI/OpenMP
is the actual programming model in HPC \cite{EXASCALE-2010}.  Per example the performance of the ScaLAPACK library can be improved by a multithread BLAS library (MKL, ESSLSMP or ACML). 
Although this ideology gives results, the threads management  gives an additional workload, when the  grain problem size will become too fine \cite{SMITH-2001}.

@@

In this paper, we proposed an alternative approach of programming.  We developed a powerful oriented object framework named Ambient, that transposed  the CUDA ideology to a cluster. 
As in the CUDA framework, in Ambient, the developer writes kernels which are executed in parallel, dynamically, and independently if needed, all other the resource of the cluster.
The needed communications  are assured by a nested MPI layer away from the users. From the user point of view, the objective is to write  parallel program without interacting the MPI API.

In this paper, we are going to describe the difference of workflow between an Ambient  and a CUDA program,  describe the blue print of Ambient and evaluated on several  benchmarks.  
Until the end of this paper, we will apply and illustrate our framework on basic linear algebra problems.

@@

As soon as Ambient encounters the flush signal (contrary to CUDA, the programmer orders the execution   as in queue model) kernel
execution is initiated along the following sequence:

@@

When the framework encounters the signal to flush the stack of kernels the actual execution starts. The framework's engine goes through the accumulated stack 
and executes all independent operations at once. Executing an operation, the engine first invokes the associated logistic kernel and allocates a specified
amount of resources for this operation. Next, the engine sets the argument's (here: the matrices) tiling using user specified work group sizes and maps resulting work groups onto allocated
processes. Now the engine can freely transfer the tiled data from previous owners to the allocated processes changing
the tile size if needed. After the transfer is complete the computational kernel kicks in behaving almost the same way as in CUDA, with work groups
simply being mapped onto threads of MPI process as if they were GPU SMP (the difference lies mainly in the user's control over the mapping of processes to work groups).

@@

A CPU core has much higher capabilities than a GPU core, and thus we can't simply downcast the CPU core to the GPU one.
Therefore, the scope of the data for a single CPU (on which a single Ambient thread typically runs) in Ambient is rather comparable to the capacity of CUDA's thread block or an OpenCL work group. The typical block size is $ 128 \times 128$, although, the user has full control of this size and can adapt it to the needs of the target problem. As in CUDA, work groups and corresponding grid of groups are formed in Ambient. A basic comparison between CUDA and Ambient is shown in table \ref{ANALOGY}.

@@

The work group approach imposes the data layout of the framework. The standard math libraries impose both column order as well as a contiguous memory order on a matrix storage (figure \ref{MEMMATRIX}).

@@

In CUDA and OpenCL, the execution model imposes kernel initialization with the appropriate resources (\eg block and grid dimensions),
after which the user does not have to explicitely manage the memory transfer inside the GPU; due to the distributed memory of clusters and the frequently heterogeneous nature of processing units we need both fine-grained control of the work group per process assignement as well as of the memory transfer operations. That should be done in a simple, automatic yet flexible way. In order to achieve this goal we split the kernel invocation into two parts.

@@

Next, we generalize example (\ref{3OPERATIONS}) to deal with vectors of matrices. We assume conditional statements determining which matrices should be added and/or multipled (Table \ref{GRAPH}). First, from line 1 to 11, as a dry run, the application execution will determine which operations should be performed, 
pushing the required kernels on the top of the stack. When the \texttt{playout()} function is called in line 12, the engine will execute the stack. By resolving dependencies it will naturally create the directed acyclic graph of the application. 
\begin{table}[t] 
	\begin{center}
	 	\begin{tabular}{l l }
                           \hline 
                           Determination of the directed  acyclic graph   &    \\ \hline
                           \textbf{Require}    A class Matrix  that adheres Ambient architecture   &  \\  
                           and  vector of matrices  of size $n$, $A[n]$, $B[n]$, $C[n]$ and $D[n]$ & \\
                           \begin{tabular}{c l} 
                               \tiny{1:}   & \textbf{for} from $i$  to the size of the vector \textbf{do} \\
                               \tiny{2:}   & \hspace{0.2 cm} \textbf{for} from $j$  to the size of the vector \textbf{do} \\                               
                               \tiny{3:}   & \hspace{0.4 cm}  \textbf{if}  the conditions on $A[i]$ and  $B[j]$ are verified \textbf{do} \\                               
                               \tiny{4:}   & \hspace{0.8 cm}  $A[i]+ \!\! =B[j]$, $\vartriangleright$  push l/c kernel 1, \\
                               \tiny{5:}   & \hspace{0.4 cm} \textbf{end if} \\   
                               \tiny{6:}   & \hspace{0.4 cm} $A[i]\times \!\! =\lambda$,  $\vartriangleright$  push l/c kernel 2, \\                   
                               \tiny{7:}   & \hspace{0.4 cm} \textbf{if}  the conditions on $C[i]$ and  $D[j]$ are verified \textbf{do} \\                               
                               \tiny{8:}   & \hspace{0.8 cm} $C[i]\times \!\! =D[j]$,  $\vartriangleright$ push l/c kernel 3, \\
                               \tiny{9:}   & \hspace{0.4 cm}  \textbf{end if} \\       
                               \tiny{10:} & \hspace{0.2 cm}  \textbf{end do} \\   
                               \tiny{11:} & \textbf{end do} \\                                 
                               \tiny{12:} & \texttt{playout()} $\vartriangleright$  execute the stack \\
                               \tiny{13:} & $\vartriangleright$ Note:  $+\!\!=$ and $\times \!\! =$ operators have wrapper functions   \\
                                                & that push the good logistic and computational kernel. \\                               
                           \end{tabular}   &  \\ \hline
        	     \end{tabular}
	\end{center}
	\caption{Generalization of the example \ref{3OPERATIONS} with conditional statements. The cumulativeness of the stack and delayed manner of execution compromise the directed acyclic graph together. \label{GRAPH}}
\end{table}

@@

First is the object's lifetime. As we employ the technique of accumulated execution, we have to ensure that the objects that were created on the stack are still alive at the time of the actual execution. For this purpose we introduced \texttt { livelong } class that forces derived data types to duplicate themselves in heap memory upon construction if they were allocated on stack. Using this duplicate we can ensure that the object will stay alive until the moment of execution of the actual kernels using this data.

>>>>

In the last decade, the distributed memory clusters were confirmed like the technology in HPC architecture machine, 100 \% of the top 500 are now based on this model.  The large scalability applications
used  a main MPI layer, we can cite as popular   in a few scientist field :  [GIVE ME A MPI CODE INTO ALPS !!!!]   {\sc Quantum ESPRESSO} \cite{QE-2009} in quantum chemistry, Abinit \cite{ABINIT-2005, ABINIT-2009} in 
quantum mechanic,   ScaLAPACK \cite{SCALAPACK-1997} in linear  algebra or SMILE in hypersonic rarefied flow \cite{SMILE-2005}. Of course, this  list is not exhaustive, and it can be fill up easily in every fields of 
science and engineering. The limitation of this programming model (the scalability) will be the communication overhead and the workload imbalance among process.

Secondly, the multicore technology  introduced by the ship makers the last decade.  Indeed, the progress of the industrial process allowed the integration of several cores into the same die. 
This technology is now available on all  processor architectures (Power, X86, ARM, \etc). One more time, the programming 
model was modified, and it re-actualized the shared memory model.  To program shared memory machine, a proved technology exists, it is the POSIX threads library. 
The programmers have two possibilities. They utilize the library directly (pThreads) or they let the compiler generates a multi-thread executables.  
Generally, the former option places the burden on the programmer, but when well done, it provides good performance.
On the other hand,  the second option more knew under the  name OpenMP  \cite{openmp:2002}, easier,  is based on  preprocessor directive. This last alternative
is clearly the most popular nowadays.

As MPI, OpenMP is developed into a consortium, the version 1.0 was released in 1997, the 2.0 in 2005 and new features were integrated in 2008.
It is now integrated natively in all compilers and it allows hybrid programs. Hybrid mode MPI/OpenMP
is the actual programming model in HPC \cite{EXASCALE-2010}.  Per example the performance of the ScaLAPACK library can be improved by a multithread BLAS library (MKL, ESSLSMP or ACML). 
Although this ideology gives results, the threads management  gives an additional workload, when the  grain problem size will become too fine \cite{SMITH-2001}.

@@

The purpose of the  paper is to present a new programming model based on a powerful oriented object (C++) framework named Ambient, that transposed  the CUDA ideology to a cluster. 
As in the CUDA framework, in Ambient, the developer writes kernels which are executed in parallel under MPI group  all other the resource of the cluster, dynamically, and independently if needed.
The  communications  are assured by a nested MPI layer away from the users. Our main objective is to allow parallel programming for the user without interacting  with the MPI API.

In this paper, we are going to describe the difference of workflow between an Ambient  and a CUDA program,  describe the general blue print of Ambient and evaluated on several  benchmarks on basic linear algebra.  
Until the end of this paper, we will apply and illustrate our framework on basic linear algebra problems, especially on pBLAS operations.

@@

As soon as Ambient encounters the flush signal (contrary to CUDA, the programmer decides  the execution start  as in the queue model programming)
kernel (a logistic kernel prepares the memory, a computational kernel execute pure calculation,   a deeper description of the pair logistics/computational kernels is given in the next section) execution is initiated along the following sequence:

@@

The execution of kernels named \guillemotleft  accumulated execution \guillemotright is one of the most important purpose of Ambient because it can be parallel.
Indeed several kernels can be push into the stack before the execution signal. When the framework encounters the signal to flush the stack of kernels the actual execution starts. 
The framework's engine goes through the accumulated stack  and executes all independent operations at once. Executing an operation, the engine first invokes the associated logistic kernel and allocates a specified
amount of resources for this operation. Next, the engine sets the argument's (here: the matrices) tiling using user specified work group sizes and maps resulting work groups onto allocated
processes. Now the engine can freely transfer the tiled data from previous owners to the allocated processes changing
the tile size if needed. After the transfer is complete the computational kernel kicks in behaving almost the same way as in CUDA, with work groups
simply being mapped onto threads of MPI process as if they were GPU SMP (the difference lies mainly in the user's control over the mapping of processes to work groups).

@@

A CPU core has much higher capabilities than a GPU core, and thus we can't simply downcast the CPU core to the GPU one.
Therefore, the scope of the data for a single CPU (on which a single Ambient thread typically runs) in Ambient is rather comparable to the capacity of CUDA's thread block or an OpenCL work group. The typical block size is $ 128 \times 128$, although, the user has full control of this size and can adapt it to the needs of the target problem. As in CUDA, work groups and corresponding grid of groups are formed in Ambient. A basic comparison between CUDA and Ambient is shown in table \ref{ANALOGY}.
Moreover the representation of the data by work group fits very well with linear algebra and the matrixes representations (basic types).

@@

The work group approach imposes the data layout of the framework. The standard math libraries impose both column-major-order as well as a contiguous memory order on a matrix storage (figure \ref{MEMMATRIX}).

@@

In CUDA and OpenCL, the execution model imposes kernel initialization with the appropriate resources (\eg block and grid dimensions),
after which the user does not have to explicitly manage the memory transfer inside the GPU; due to the distributed memory of clusters and the frequently heterogeneous nature of processing units we need both fine-grained control of the work group per process assignement as well as of the memory transfer operations. That should be done in a simple, automatic yet flexible way. In order to achieve this goal we split the kernel invocation into two parts.

@@

Next, we generalize example (\ref{3OPERATIONS}) to deal with vectors of matrices. We assume conditional statements determining which matrices should be added and/or multipled (Table \ref{GRAPH}). First, from line 1 to 11, as a dry run, the application execution will determine which operations should be performed, 
pushing the required kernels on the top of the stack. When the \texttt{playout()} function is called in line 12, the engine will execute the stack. By resolving dependencies it will naturally create the directed acyclic graph of the application. 
\begin{table}[t] 
	\begin{center}
	 	\begin{tabular}{l l }
                           \hline 
                           Determination of the directed  acyclic graph   &    \\ \hline
                           \textbf{Require}    A class Matrix  that adheres Ambient architecture   &  \\  
                           and  vector of matrices  of size $n$, $A[n]$, $B[n]$, $C[n]$ and $D[n]$ & \\
                           \begin{tabular}{c l} 
                               \tiny{1:}   & \textbf{for} from $i$  to the size of the vector \textbf{do} \\
                               \tiny{2:}   & \hspace{0.2 cm} \textbf{for} from $j(0)$  to the size of the vector \textbf{do} \\                               
                               \tiny{3:}   & \hspace{0.4 cm}  \textbf{if}  the conditions on $A[i]$ and  $B[j]$ are verified \textbf{do} \\                               
                               \tiny{4:}   & \hspace{0.8 cm}  $A[i]+ \!\! =B[j]$, $\vartriangleright$  push l/c kernel 1, \\
                               \tiny{5:}   & \hspace{0.4 cm} \textbf{end if} \\   
                               \tiny{6:}   & \hspace{0.4 cm} $A[i]\times \!\! =\lambda$,  $\vartriangleright$  push l/c kernel 2, \\                   
                               \tiny{7:}   & \hspace{0.4 cm} \textbf{if}  the conditions on $C[i]$ and  $D[j]$ are verified \textbf{do} \\                               
                               \tiny{8:}   & \hspace{0.8 cm} $C[i]\times \!\! =D[j]$,  $\vartriangleright$ push l/c kernel 3, \\
                               \tiny{9:}   & \hspace{0.4 cm}  \textbf{end if} \\       
                               \tiny{10:} & \hspace{0.2 cm}  \textbf{end do} \\   
                               \tiny{11:} & \textbf{end do} \\                                 
                               \tiny{12:} & \texttt{playout()} $\vartriangleright$  execute the stack \\
                               \tiny{13:} & $\vartriangleright$ Note:  $+\!\!=$ and $\times \!\! =$ operators have wrapper functions   \\
                                                & that push the good logistic and computational kernel. \\                               
                           \end{tabular}   &  \\ \hline
        	     \end{tabular}
	\end{center}
	\caption{Generalization of the example \ref{3OPERATIONS} with conditional statements. The cumulativeness of the stack and delayed manner of execution compromise the directed acyclic graph together. \label{GRAPH}}
\end{table}

@@

First is the object's lifetime. Due to the accumulated execution model, objects may be used after their lexical scope has expired.
For this purpose we introduced \texttt { livelong } class that forces derived data types to duplicate themselves in heap memory upon construction if they were allocated on stack. Using this duplicate we can ensure that the object will stay alive until the moment of execution of the actual kernels using this data.
